---
title: "P597-6_Elucidating_the_Function_of_anti-avb6_autoantibodies_in_UC"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
editor_options: 
  chunk_output_type: console
date: "2025-06-24"
---


# Project Description

[description goes here]

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, results = "hide"}
library(knitr)
opts_chunk$set(
  fig.width = 6, fig.height = 4.25, cache = TRUE,
  echo = FALSE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
library(dplyr)
library(ggplot2)
library(ggVennDiagram) # for making the diagram
library(gplots) # for retrieving the lists
library(lintr)
# linters <- linters_with_defaults(
#   assignment_linter = NULL,
#   line_length_linter(120)
# )
# lint(filename, linters = linters)
theme_set(
  theme_bw(base_size = 14) +
    theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, size = 1),
      axis.text = element_text(colour = "black"),
      axis.ticks = element_line(colour = "black"),
      legend.key = element_blank(),
      strip.text.x = element_text(size = 14, margin = margin(b = 2, t = 2)),
      strip.background = element_rect(fill = "white", colour = "black")))
library(ggrepel)
options(ggrepel.max.overlaps = Inf)
library(miscHelpers)
library(annotables)
library(apird)
library(magrittr)
library(readxl)
library(tidyverse)
# library(Seurat)
# library(SeuratWrappers)
# library(monocle3)
library(ggbeeswarm)
library(stringr)
library(kableExtra)
library(gtools)
library(inlmisc) # For colors
library(data.table)
library(RNAseQC)
# library(SingleCellExperiment)
library(plotscale)
library(gtable)
library(grid)
library(ggh4x)
# library(concaveman)
# library(ggforce)
# library(gghighlight)
library(ComplexHeatmap)
library(limma)
library(edgeR)
library(caret)
library(scales)
# library(CDI) # this is a development version
# library(clusterSim) # for clustering quality evaluation functions (DB, CH, etc)
library(tictoc)
library(foreach)
library(doParallel)
library(viridis)
library(viridisLite)
library(GGally)
library(RColorBrewer)
library(pals)
library(plotly)
library(ggpmisc)
library(haven)
library(circlize)
library(ggvenn)
library(ggnewscale)
library(msigdbr)
library(clusterProfiler)
library(enrichplot)
library(openxlsx)
library(paletteer)
library(genekitr)
```

```{r set_up_directories, cache = TRUE}
baseDir <- "/Users/tedwards/Documents/projects/P597-6_Elucidating_the_Function_of_anti-avb6_autoantibodies_in_UC"
outputDataDir <- file.path(baseDir, "data/outputData")
inputDataDir <- file.path(baseDir, "data/inputData")
plotDir <- file.path(baseDir, "figures")

dataDate <- "2025-06-30" # yyyy-mm-dd

filenameSuffix <- paste0("P597_6_", dataDate)
```

```{r setupFunctions}
# function for saving plots as both pdf and png
savePlot <- function(
    plot,
    plotDir,
    filename,
    height,
    width,
    units = "in",
    dpi = 600,
    formats = c("pdf", "png")
    ) {
  # Ensure plotDir exists
  if (!dir.exists(plotDir)) dir.create(plotDir, recursive = TRUE)

  # Save as PDF
  if ("pdf" %in% formats) {
    pdf(file.path(plotDir, paste0(filename, ".pdf")), height = height, width = width)
    print(plot)
    dev.off()
  }

  # Save as PNG
  if ("png" %in% formats) {
    png(
      file.path(plotDir, paste0(filename, ".png")),
      height = height,
      width = width,
      units = units,
      res = dpi
    )
    print(plot)
    dev.off()
  }
}
```

```{r import_data}
# load the sequencing data
# Use apird function to get project data
# project <- "P597-6"

# # getGcqProjectInfo(project)
# libids <- getProjectLibs(project, searchType = "regex")

# # retrieve data
# anno <- getAnno(libids)
# metrics <- getMetrics(libids)
# counts <- getGeneCounts(libids)

# # Save for offline analysis
# save(anno,
#   metrics,
#   counts,
#   file = file.path(inputDataDir, "P597_6_Data.Rdata"))

load(file.path(inputDataDir, "P597_6_Data.Rdata"))


# load in gene modules
# geneModules <-
#   read_excel(file.path(inputDataDir, "P597_geneModules.xlsx")) %>%
#   data.frame()
```

```{r cleanGenomicsData}
# fix column names
anno <- anno %>%
  dplyr::rename(sampleName = sample_name,
    sampleId = sample_id)


# replace spaces with underscores, remove hyphens
anno$sampleName <- gsub("-", "", gsub("[[:space:]]+", "_", anno$sampleName))
anno$treatment <- gsub("-", "", gsub("[[:space:]]+", "_", anno$treatment))
anno$project <- gsub("-", "", gsub("[[:space:]]+", "_", anno$project))
anno$cellType <- gsub("-", "", gsub("[[:space:]]+", "_", anno$cellType))

# replace "TNFa_TGFb" with "TNFa_antiTGFb" in anno$treatment
anno$treatment <- gsub("TNFa_TGFb", "TNFa_antiTGFb", anno$treatment)

# make factors
anno$treatment <- factor(anno$treatment)

# make $stimulation where $IL4* maps to "IL4", "IL4_IL17*" maps to "IL4_IL17", "TNFa*" maps to "TNFa", "TNFa_IL17*" maps to "TNFa_IL17", and "NoTx" is "noStim",
# accounting for the conflicts between the IL17/non-IL17 stims
anno$stimulation <- case_when(
  str_detect(anno$treatment, "^IL4_IL17") ~ "IL4_IL17",
  str_detect(anno$treatment, "^TNFa_IL17") ~ "TNFa_IL17",
  str_detect(anno$treatment, "^IL4") ~ "IL4",
  str_detect(anno$treatment, "^TNFa") ~ "TNFa",
  anno$treatment == "NoTx" ~ "noStim",
  TRUE ~ as.character(anno$treatment)  # fallback for any other treatments
)

# join annotations with metrics
annotatedMetrics <- dplyr::full_join(anno, metrics, by = c("libid"))

# transpose counts matrix
counts <- t(counts)

# fix the libids
colnames(counts) <- sub("_.*", "", colnames(counts))
```

```{r setupPalettes}

colors_sel <- c(
  "#053061",  # Dark blue
  "#2166AC",  # Medium blue
  "#4393C3",  # Light blue
  "#92C5DE",  # Very light blue
  "#D1E5F0",  # Pale blue
  "#FFFFFF",  # White (center/neutral)
  "#FDDBC7",  # Pale red
  "#F4A582",  # Very light red
  "#D6604D",  # Light red
  "#B2182B",  # Medium red
  "#67001F"   # Dark red
)


# treatment

# colors from palettesForR::Tango
palTreatment <- c("IL4" = "#FCAF3EFF",
  "IL4_3G9" = "#F57900FF",
  "IL4_antiTGFb" = "#CE5C00FF",
  "IL4_IL17" = "#729FCFFF",
  "IL4_IL17_3G9" = "#3465A4FF",
  "IL4_IL17_antiTGFb" = "#204A87FF",
  "TNFa" = "#AD7FA8FF",
  "TNFa_3G9" = "#75507BFF",
  "TNFa_antiTGFb" = "#5C3566FF",
  "TNFa_IL17" = "#EF2929FF",
  "TNFa_IL17_3G9" = "#CC0000FF",
  "TNFa_IL17_antiTGFb" = "#A40000FF",
  "NoTx" = "grey")

palStimulation <- c("IL4" = "#F57900FF",
  "IL4_IL17" = "#3465A4FF",
  "TNFa" = "#75507BFF",
  "TNFa_IL17" = "#CC0000FF",
  "noStim" = "grey")
```


```{r setQCCuts}
# Set QC cuts
alignCut <- 80 # 70
totalReadsCut <- 0.5
medianCVCut <- 0.8

annotatedMetrics$qcPass <- annotatedMetrics$fastq_total_reads > totalReadsCut * 10^6 & # require more than 500,000 total reads
  annotatedMetrics$pct_aligned > alignCut & # require 80 percent alignment or better
  annotatedMetrics$median_cv_coverage < medianCVCut
```


```{r qcplots_total_reads, fig.width=7, fig.height=3.5}
# Make a barplot of total reads. Arrange libraries from lowest number of reads to highest.
gTotReads.tmp <- ggplot(data = annotatedMetrics,
  aes(x = reorder(libid, fastq_total_reads),
    y = fastq_total_reads / 10^6,
    fill = treatment)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = totalReadsCut) +
  scale_fill_manual(values = palTreatment) +
  labs(x = "Library", y = "Total reads in millions", fill = "") +
  theme(text = element_text(size = 12)) +
  theme(axis.text.x = element_blank(),
    axis.ticks.x = element_blank()) #+
#   facet_wrap(~cellType)

pdf(
  gTotReads.tmp,
  file = file.path(plotDir,
    paste0(dataDate, "_TotalReads.pdf")),
  height = 4,
  width = 10
)

print(gTotReads.tmp)

invisible(dev.off())
rm_tmp(ask = FALSE)
```

```{r qc_cut_plot, fig.width=6, fig.height=4}

annotatedMetrics$median_cv_coverage <- as.numeric(annotatedMetrics$median_cv_coverage)

gPctAligned.tmp <- ggplot(data = annotatedMetrics) +
  geom_point(aes(x = median_cv_coverage,
    y = pct_aligned,
    color = treatment),
  size = 1.5) +
  scale_color_manual(values = palTreatment) +
  labs(x = "Median CV coverage",
    y = "Percent alignment",
    color = "") +
  geom_hline(yintercept = alignCut) +
  geom_vline(xintercept = medianCVCut) +
  ylim(c(0, 100)) +
  theme(text = element_text(size = 12))

pdf(
  gPctAligned.tmp,
  file = file.path(plotDir,
    paste0(dataDate, "MedianCVvsPctAligned.pdf")),
  height = 5,
  width = 7
)
print(gPctAligned.tmp)

invisible(dev.off())
rm_tmp(ask = FALSE)
```

```{r make_qc_cuts}
annotatedMetricsQC <- annotatedMetrics %>%
  dplyr::filter(qcPass == TRUE)

countsQC <- counts[, colnames(counts) %in% annotatedMetricsQC$libid]

# Make sure libraries in counts and metrics are in the same order
libOrder <- match(colnames(countsQC), annotatedMetricsQC$libid)
annotatedMetricsQC <- annotatedMetricsQC[libOrder, ]

pctPassQC <- round(nrow(annotatedMetricsQC) / nrow(annotatedMetrics) * 100)
```

```{r qc_failures, results = "asis"}
failedSamples <- annotatedMetrics %>%
  dplyr::filter(qcPass == FALSE) %>%
  dplyr::select(libid,
    sampleName) %>%
  dplyr::arrange(libid,
    sampleName)

failedSamples %>%
  kable(row.names = F) %>%
  kable_styling("striped",
    full_width = F,
    position = "left") %>%
  scroll_box(height = "800px")
```

```{r gene_filtering}
## Keep protein coding genes with HGNC symbols, and drop non-protein-coding genes

## This chunk avoids an issue with the annotables update. consider cleaning up/elegantizing.
v38 <- annotables::grch38
pc38 <- v38[v38$biotype == "protein_coding", ] # grab protein coding
pcWithNoSymbol <- pc38[pc38$symbol == "", ] # protein coding that have no symbol
pcWithSymbol <- pc38[pc38$symbol != "", ] # get the protein coding entries that have symbols
countsWithSymbol <- countsQC[rownames(countsQC) %in% pcWithSymbol$ensgene, ] # make a counts matrix that is protein coding and has symbol

countsTmp <- countsQC[rownames(countsQC) %in% pcWithSymbol$ensgene, ]
countsTmp <- as.data.frame(countsTmp)
countsTmp$gene <- pcWithSymbol$symbol[match(rownames(countsTmp), pcWithSymbol$ensgene)]
countsTmp <- as.data.table(countsTmp)

countsPC <-
  countsTmp[
    , lapply(.SD, sum), by = gene, .SDcols = grep("^lib", colnames(countsTmp), value = TRUE)] %>%
  arrange(gene) %>%
  dplyr::filter(gene != "MTRNR2L1") %>% # exclude MTRNR2L1 # THE I don't yet know why we'd exclude a gene TODO understand
  as.data.frame() %>%
  magrittr::set_rownames(., value = .$gene)

countsPC <- countsPC[, -which(colnames(countsPC) == "gene")]

# dim(counts_pc)
## filter lowly expressed genes

# Define a function to filter out lowly expressed genes at least 1 cpm in 10% of libraries (default)
gene_filter <- function(counts_in,
                        per_cutoff = 0.1,
                        counts_cutoff = 1) {
  # Keep genes with cpm of at least counts_cutoff in at least per_cutoff fraction of libraries
  # CPM normalize
  counts_cpm_norm <- as.data.frame(t(t(counts_in * 10^6) / colSums(counts_in)))

  # Filter out lowly expressed genes
  keepRows <- rowSums((counts_cpm_norm) >= counts_cutoff) >= per_cutoff * ncol(counts_cpm_norm)
  counts_filtered <- counts_in[keepRows, ]

  return(counts_filtered)

}

# Run function to filter lowly expressed genes
# counts_all_filtered <- gene_filter(counts_hgnc, 0.10) #all genes with HGNC symbols
countsPCFiltered <- gene_filter(countsPC, 0.1, 1)

normalize_counts <- function(counts_in, method) {
  # normalize using tmm or deconvolution
  # tmm is good for bulk RNAseq
  # deconvolution is best for large datasets of single cell RNAseq
  # deconvolution is NOT recommended for smaller datasets (less than a few hundred cells)

  if (method == "decon") {
    # Normalize using the deconvolution algorithm
    # Normalize using the deconvolution algorithm
    sce <- SingleCellExperiment(assays = SimpleList(counts = as.matrix(counts_in)))
    sce <- scran::computeSumFactors(sce)
    sce <- scater::logNormCounts(sce)
    # counts_norm <- as.data.frame(t(t(counts_in)/decon_norm_factors))
    counts <- assay(sce)
    libsizes <- colSums(counts)
    size.factors <- libsizes / mean(libsizes)
    counts_norm <- as.data.frame(t(t(counts) / size.factors))

    # Deprecated approach. Does not work in R version >4
    # decon_norm_factors <- computeSumFactors(as.matrix(counts_in))
    # counts_norm <- as.data.frame(t(t(counts_in)/decon_norm_factors))
  }

  if (method == "tmm") {
    # Normalize using the TMM algorithm
    dge <- DGEList(counts_in)
    dge <- calcNormFactors(dge)
    counts_norm <- edgeR::cpm(dge, normalized.lib.sizes = TRUE)
  }

  return(counts_norm)

}


countsPCNorm <- normalize_counts(countsPCFiltered, "tmm")
# counts_all_norm <- normalize_counts(counts_all_filtered, "tmm")

# Create a dataframe of gene names to attach to the DGEList(). Keep both ensembl_gene_id, and hgnc_symbol
fgenes  <- pcWithSymbol[match(rownames(countsPCNorm), pcWithSymbol$ensgene), c("ensgene", "symbol")]
dge <- DGEList(counts = countsPCNorm, genes = fgenes)

# Write out normalized counts
# Add HGNC symbols to gene names
# log 2 transform

countsOut <- log2(countsPCNorm + 1)

write.csv(countsOut, file.path(outputDataDir, "Deconvolution_normalized_log2_transformed_counts.csv"), quote = FALSE, row.names = TRUE)
```

```{r saveDataAfterQC}
# save workspace as an .Rdata file
save.image(file = file.path(outputDataDir, "workspace_after_QC.Rdata"))
```

```{r pca_all, dependson="pca_plotting_func", fig.width=9, fig.height=4}
png_res <- 600

# Run PCA on the normalized log2 transformed counts data
dat <- log2(as.data.frame(t(dge$counts)) + 0.5)
pca <- prcomp(log2(as.data.frame(t(countsPCNorm)) + 1), center = TRUE, scale = FALSE)

# Compute correlations between PCs and annotatedMetrics variables
pcCors <- calc_PCcors(pca, annotatedMetricsQC, id_col = "libid") %>%
  t()
sum_pca <- summary(pca)
pca_scores <- as.data.frame(pca$x)
pdatscores <- merge(annotatedMetricsQC, pca_scores, by.x = "libid", by.y = "row.names")
pc1_lab <- paste("PC1 (", round(100 * sum_pca$importance[2, 1], 1),  "%)", sep = "")
pc2_lab <- paste("PC2 (", round(100 * sum_pca$importance[2, 2], 1),  "%)", sep = "")
pc3_lab <- paste("PC3 (", round(100 * sum_pca$importance[2, 3], 1),  "%)", sep = "")
pc4_lab <- paste("PC4 (", round(100 * sum_pca$importance[2, 4], 1),  "%)", sep = "")
pc5_lab <- paste("PC5 (", round(100 * sum_pca$importance[2, 5], 1),  "%)", sep = "")
pc6_lab <- paste("PC6 (", round(100 * sum_pca$importance[2, 6], 1),  "%)", sep = "")

png(filename = file.path(plotDir, paste0("PCA1v2_treatment.", filenameSuffix, ".png")),
  width = 6, height = 5, units = "in", res = 300)

plot.tmp <- ggplot(pdatscores, aes(x = PC1, y = PC2, color = treatment)) +
  geom_point(size = 3) +
  scale_color_manual(values = palTreatment) +
  labs(color = "treatment", x = pc1_lab, y = pc2_lab)

print(plot.tmp)
dev.off()
rm_tmp(ask = FALSE)

png(filename =
  file.path(
    plotDir,
    paste0("PCA2v3_treatment.", filenameSuffix, ".png")),
width = 6, height = 5,
units = "in",
res = 300)
plot.tmp <- ggplot(pdatscores, aes(x = PC2, y = PC3, color = treatment)) +
  geom_point(size = 3) +
  scale_color_manual(values = palTreatment) +
  labs(color = "cellType") +
  labs(x = pc2_lab, y = pc3_lab)
print(plot.tmp)
dev.off()
rm_tmp(ask = FALSE)

png(filename =
  file.path(
    plotDir,
    paste0("PCA3v4_treatment.", filenameSuffix, ".png")),
width = 6, height = 5,
units = "in",
res = 300)
plot.tmp <- ggplot(pdatscores, aes(x = PC3, y = PC4, color = treatment)) +
  geom_point(size = 3) +
  scale_color_manual(values = palTreatment) +
  labs(color = "cellType") +
  labs(x = pc2_lab, y = pc3_lab)
print(plot.tmp)
dev.off()
rm_tmp(ask = FALSE)
```

```{r diffexp_treatment_allSubjects_donor_rand_effect}
countsPCNorm  <- countsPCNorm[, annotatedMetricsQC$libid]

designMat.treatment <- model.matrix(~ 0 + treatment, data = annotatedMetricsQC)
colnames(designMat.treatment)  <- stringr::str_remove(colnames(designMat.treatment), "treatment")

print(colnames(designMat.treatment))
colnames(designMat.treatment) <- make.names(colnames(designMat.treatment))

vwts.treatment  <- voom(countsPCNorm, design = designMat.treatment, plot = TRUE, span = 0.1)

# cor <- duplicateCorrelation(vwts.treatment, designMat.treatment, block = annotatedMetricsQC$libid)

vwts.treatment  <- voom(countsPCNorm,
  design = designMat.treatment,
  # correlation = cor$consensus.correlation,
  plot = TRUE, span = 0.1)

fit.treatment <- lmFit(vwts.treatment,
  design = designMat.treatment,
  # block = annotatedMetricsQC$libid,
  # correlation = cor$consensus.correlation
)

fit.treatment <- eBayes(fit.treatment)

contrasts.matrix <- makeContrasts(
  `IL4_vs_NoTx`                     = `IL4` - `NoTx`,
  `IL4_3G9_vs_NoTx`                 = `IL4_3G9` - `NoTx`,
  `IL4_antiTGFb_vs_NoTx`            = `IL4_antiTGFb` - `NoTx`,

  `IL4_IL17_vs_NoTx`                = `IL4_IL17` - `NoTx`,
  `IL4_IL17_3G9_vs_NoTx`            = `IL4_IL17_3G9` - `NoTx`,
  `IL4_IL17_antiTGFb_vs_NoTx`       = `IL4_IL17_antiTGFb` - `NoTx`,

  `TNFa_vs_NoTx`                    = `TNFa` - `NoTx`,
  `TNFa_3G9_vs_NoTx`                = `TNFa_3G9` - `NoTx`,
  `TNFa_antiTGFb_vs_NoTx`           = `TNFa_antiTGFb` - `NoTx`,

  `TNFa_IL17_vs_NoTx`               = `TNFa_IL17` - `NoTx`,
  `TNFa_IL17_3G9_vs_NoTx`           = `TNFa_IL17_3G9` - `NoTx`,
  `TNFa_IL17_antiTGFb_vs_NoTx`      = `TNFa_IL17_antiTGFb` - `NoTx`,

  `IL4_3G9_vs_IL4`                  = `IL4_3G9` - `IL4`,
  `IL4_antiTGFb_vs_IL4`             = `IL4_antiTGFb` - `IL4`,

  `IL4_IL17_3G9_vs_IL4_IL17`        = `IL4_IL17_3G9` - `IL4_IL17`,
  `IL4_IL17_antiTGFb_vs_IL4_IL17`   = `IL4_IL17_antiTGFb` - `IL4_IL17`,

  `TNFa_3G9_vs_TNFa`                = `TNFa_3G9` - `TNFa`,
  `TNFa_antiTGFb_vs_TNFa`           = `TNFa_antiTGFb` - `TNFa`,

  `TNFa_IL17_3G9_vs_TNFa_IL17`      = `TNFa_IL17_3G9` - `TNFa_IL17`,
  `TNFa_IL17_antiTGFb_vs_TNFa_IL17` = `TNFa_IL17_antiTGFb` - `TNFa_IL17`,

  levels = designMat.treatment)

fit.treatment_contrasts <- contrasts.fit(fit.treatment, contrasts.matrix)

vfit.treatment_contrasts  <- eBayes(fit.treatment_contrasts, robust = TRUE)

# topTableBaselineVsScreen <- topTable(vfit.treatment_contrasts, coef = "baselineVsScreen", number = Inf, sort.by = "P")
# topTableBaselineVsScreen$gene <- rownames(topTableBaselineVsScreen) # get genenames from rownames

# p_cutoff <- 0.05
# logfc_cutoff <- 1
# sig.genes <- topTableBaselineVsScreen$adj.P.Val <= p_cutoff
# print(table(sig.genes))

comparisons <- list()
for (i in 1:ncol(contrasts.matrix)) {
  compname <- colnames(contrasts.matrix)[i]
  z <- topTable (vfit.treatment_contrasts, coef = i, number = Inf, sort.by = "P")
  z$gene <- rownames(z)
  comparisons[[compname]] <- z
  print(compname)
  sigs <- z$adj.P.Val <= 0.05
  print(table(sigs))
  write.csv(z, file = file.path(outputDataDir, paste0(compname, "_DE.csv")), row.names = TRUE, quote = FALSE)
}

list2env(comparisons, envir = .GlobalEnv)

# print a table containing the number of significant DE genes in each contrast, split up by positive/negative logFC
# Create a table of significant DE genes for each contrast
sig_table <- do.call(rbind, lapply(names(comparisons), function(comp) {
  df <- comparisons[[comp]]
  sig <- df$adj.P.Val <= 0.05
  up <- sum(sig & df$logFC > 0, na.rm = TRUE)
  down <- sum(sig & df$logFC < 0, na.rm = TRUE)
  tot <- up + down
  data.frame(Contrast = comp,
    Up = up,
    Down = down,
    Total = tot,
    stringsAsFactors = FALSE)
}))

# Print the table
print(sig_table)

# save sig_table as a .csv
write.csv(sig_table, file = file.path(outputDataDir, "sig_table.csv"), row.names = FALSE)
```

```{r makeVolcanoPlots}
# Create a list to store the volcano plots (if needed later)
volcano_plots <- list()

# Loop over each contrast in your comparisons list
for (comp in names(comparisons)) {
  df <- comparisons[[comp]]

  # Compute -log10(P.Value)
  df <- df %>% mutate(negLogP = -log10(adj.P.Val))

  # Select top 30 genes by smallest adjusted P value
  top30 <- df %>% arrange(adj.P.Val) %>% slice(1:30)
  # Only keep those with adj.P.Val <= 0.05
  top30_sig <- top30 %>% filter(adj.P.Val <= 0.05)

  p <- ggplot(df, aes(x = logFC, y = negLogP)) +
    geom_point(alpha = 0.5) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey") +
    # Only label if there are any significant genes
    {
      if (nrow(top30_sig) > 0) geom_text_repel(data = top30_sig, aes(label = gene), size = 3, max.overlaps = Inf) else NULL
    } +
    labs(title = comp,
      x = "Log2 Fold Change",
      y = "-log10(p-value)") +
    theme_bw()

  volcano_plots[[comp]] <- p

  # Use pdf() to output the plot
  pdf_filename <- file.path(plotDir, paste0(comp, "_volcano.pdf"))
  pdf(file = pdf_filename, width = 6, height = 5)
  print(p)
  dev.off()
}
```

```{r twoGroupDEGHeatmaps}
# z-scored log expression heatmaps of sig. DEGs from the two-group comparisons

library(ComplexHeatmap)
library(circlize)

# Get list of all two-group comparisons (exclude the modifier effect comparisons)
two_group_comparisons <- names(comparisons)

# Function to get treatment groups from comparison name
get_treatment_groups <- function(comp_name) {
  # Parse comparison name to get the two treatment groups
  # Format: "treatment1_vs_treatment2"
  parts <- strsplit(comp_name, "_vs_")[[1]]
  treatment1 <- parts[1]
  treatment2 <- parts[2]
  return(list(treatment1 = treatment1, treatment2 = treatment2))
}

# Create heatmaps for each comparison
for (comp_name in two_group_comparisons) {
  cat("Creating heatmaps for:", comp_name, "\n")

  # Get the comparison data
  comp_data <- comparisons[[comp_name]]

  # Filter for significant genes
  sig_genes <- comp_data %>% filter(adj.P.Val <= 0.05)

  if (nrow(sig_genes) == 0) {
    cat("No significant genes found for", comp_name, "- skipping heatmaps\n")
    next
  }

  cat("Found", nrow(sig_genes), "significant genes for", comp_name, "\n")

  # Get treatment groups for this comparison
  treatment_groups <- get_treatment_groups(comp_name)

  # Get libraries for version b (only those involved in this comparison)
  libs_subset <- annotatedMetricsQC$libid[annotatedMetricsQC$treatment %in% c(treatment_groups$treatment1, treatment_groups$treatment2)]
  anno_subset <- annotatedMetricsQC[annotatedMetricsQC$libid %in% libs_subset, ]

  # === VERSION 1: Top 60 DEGs by adj.p.val ===
  top60_genes <- comp_data %>%
    arrange(adj.P.Val) %>%
    slice(1:min(60, nrow(comp_data))) %>%
    filter(adj.P.Val <= 0.05)  # Only include significant genes

  if (nrow(top60_genes) > 0) {
    # VERSION 1A: All libraries
    # Get expression data for these genes
    expr_data_v1a <- vwts.treatment$E[top60_genes$gene, , drop = FALSE]

    # Z-score scale by gene (row-wise)
    expr_data_v1a_scaled <- t(scale(t(expr_data_v1a)))

    # Create column annotation
    col_ann_v1a <- HeatmapAnnotation(
      Treatment = annotatedMetricsQC$treatment,
      Stimulation = annotatedMetricsQC$stimulation,
      col = list(
        Treatment = palTreatment,
        Stimulation = palStimulation
      ),
      annotation_name_gp = gpar(fontsize = 10)
    )

    # Order columns by stimulation, then by treatment
    col_order_v1a <- order(annotatedMetricsQC$stimulation, annotatedMetricsQC$treatment)
    expr_data_v1a_ordered <- expr_data_v1a_scaled[, col_order_v1a, drop = FALSE]

    # Create heatmap v1a
    # Set dynamic color range based on data
    z_range_v1a <- range(expr_data_v1a_ordered, na.rm = TRUE)
    z_max_v1a <- max(abs(z_range_v1a))

    ht_v1a <- Heatmap(
      expr_data_v1a_ordered,
      name = "Z-score",
      col = colorRamp2(c(-z_max_v1a, 0, z_max_v1a), c("blue", "white", "red")),
      show_row_names = TRUE,
      row_names_gp = gpar(fontsize = 10),
      cluster_rows = TRUE,
      column_split = annotatedMetricsQC$stimulation[col_order_v1a],
      cluster_columns = TRUE,
      cluster_column_slices = FALSE,
      show_column_names = FALSE,
      column_names_gp = gpar(fontsize = 10),
      column_names_rot = 45,
      top_annotation = col_ann_v1a,
      column_title = paste0(comp_name, ": Top ", nrow(top60_genes), " DEGs by adj.P.Val (All Libraries)"),
      column_title_gp = gpar(fontsize = 12, fontface = "bold")
    )

    # Save version 1a heatmap
    pdf_filename_v1a <- file.path(plotDir, paste0("heatmap_v1a_top60_", comp_name, ".pdf"))
    pdf(pdf_filename_v1a, width = 12, height = max(8, nrow(top60_genes) * 0.2))
    draw(ht_v1a, heatmap_legend_side = "right", annotation_legend_side = "right")
    dev.off()

    # VERSION 1B: Only comparison-relevant libraries
    expr_data_v1b <- expr_data_v1a_scaled[, libs_subset, drop = FALSE]
    expr_data_v1b_scaled <- t(scale(t(expr_data_v1b)))  # Re-scale for subset

    col_ann_v1b <- HeatmapAnnotation(
      Treatment = anno_subset$treatment,
      Stimulation = anno_subset$stimulation,
      col = list(
        Treatment = palTreatment,
        Stimulation = palStimulation
      ),
      annotation_name_gp = gpar(fontsize = 10)
    )

    col_order_v1b <- order(anno_subset$stimulation, anno_subset$treatment)
    expr_data_v1b_ordered <- expr_data_v1b_scaled[, col_order_v1b, drop = FALSE]

    # Set dynamic color range based on data
    z_range_v1b <- range(expr_data_v1b_ordered, na.rm = TRUE)
    z_max_v1b <- max(abs(z_range_v1b))

    ht_v1b <- Heatmap(
      expr_data_v1b_ordered,
      name = "Z-score",
      col = colorRamp2(c(-z_max_v1b, 0, z_max_v1b), c("blue", "white", "red")),
      show_row_names = TRUE,
      row_names_gp = gpar(fontsize = 10),
      cluster_rows = TRUE,
      column_split = anno_subset$stimulation[col_order_v1b],
      cluster_columns = TRUE,
      cluster_column_slices = FALSE,
      show_column_names = FALSE,
      column_names_gp = gpar(fontsize = 10),
      column_names_rot = 45,
      top_annotation = col_ann_v1b,
      column_title = paste0(comp_name, ": Top ", nrow(top60_genes), " DEGs by adj.P.Val (Comparison Libraries Only)"),
      column_title_gp = gpar(fontsize = 12, fontface = "bold")
    )

    pdf_filename_v1b <- file.path(plotDir, paste0("heatmap_v1b_top60_", comp_name, ".pdf"))
    pdf(pdf_filename_v1b, width = 12, height = max(8, nrow(top60_genes) * 0.2))
    draw(ht_v1b, heatmap_legend_side = "right", annotation_legend_side = "right")
    dev.off()
  }

  # === VERSION 2: Top 30 -logFC and top 30 +logFC DEGs by adj.p.val ===
  # Get top 30 downregulated genes (most negative logFC, significant)
  top30_down <- sig_genes %>%
    filter(logFC < 0) %>%
    arrange(adj.P.Val) %>%
    slice(1:min(30, nrow(.)))

  # Get top 30 upregulated genes (most positive logFC, significant)
  top30_up <- sig_genes %>%
    filter(logFC > 0) %>%
    arrange(adj.P.Val) %>%
    slice(1:min(30, nrow(.)))

  # Combine up and down genes
  top30_updown_v2 <- rbind(top30_down, top30_up)

  if (nrow(top30_updown_v2) > 0) {
    # VERSION 2A: All libraries
    expr_data_v2a <- vwts.treatment$E[top30_updown_v2$gene, , drop = FALSE]
    expr_data_v2a_scaled <- t(scale(t(expr_data_v2a)))
    expr_data_v2a_ordered <- expr_data_v2a_scaled[, col_order_v1a, drop = FALSE]

    # Set dynamic color range based on data
    z_range_v2a <- range(expr_data_v2a_ordered, na.rm = TRUE)
    z_max_v2a <- max(abs(z_range_v2a))

    ht_v2a <- Heatmap(
      expr_data_v2a_ordered,
      name = "Z-score",
      col = colorRamp2(c(-z_max_v2a, 0, z_max_v2a), c("blue", "white", "red")),
      show_row_names = TRUE,
      row_names_gp = gpar(fontsize = 10),
      cluster_rows = TRUE,
      column_split = annotatedMetricsQC$stimulation[col_order_v1a],
      cluster_columns = TRUE,
      cluster_column_slices = FALSE,
      show_column_names = FALSE,
      column_names_gp = gpar(fontsize = 10),
      column_names_rot = 45,
      top_annotation = col_ann_v1a,
      column_title = paste0(comp_name, ": Top 30 Up+Down DEGs by adj.P.Val (",
        nrow(top30_up), " up, ", nrow(top30_down), " down) - All Libraries"),
      column_title_gp = gpar(fontsize = 12, fontface = "bold")
    )

    pdf_filename_v2a <- file.path(plotDir, paste0("heatmap_v2a_top30updown_adjpval_", comp_name, ".pdf"))
    pdf(pdf_filename_v2a, width = 12, height = max(8, nrow(top30_updown_v2) * 0.25))
    draw(ht_v2a, heatmap_legend_side = "right", annotation_legend_side = "right")
    dev.off()

    # VERSION 2B: Only comparison-relevant libraries
    expr_data_v2b <- expr_data_v2a_scaled[, libs_subset, drop = FALSE]
    expr_data_v2b_scaled <- t(scale(t(expr_data_v2b)))
    expr_data_v2b_ordered <- expr_data_v2b_scaled[, col_order_v1b, drop = FALSE]

    # Set dynamic color range based on data
    z_range_v2b <- range(expr_data_v2b_ordered, na.rm = TRUE)
    z_max_v2b <- max(abs(z_range_v2b))

    ht_v2b <- Heatmap(
      expr_data_v2b_ordered,
      name = "Z-score",
      col = colorRamp2(c(-z_max_v2b, 0, z_max_v2b), c("blue", "white", "red")),
      show_row_names = TRUE,
      row_names_gp = gpar(fontsize = 10),
      cluster_rows = TRUE,
      column_split = anno_subset$stimulation[col_order_v1b],
      cluster_columns = TRUE,
      cluster_column_slices = FALSE,
      show_column_names = FALSE,
      column_names_gp = gpar(fontsize = 10),
      column_names_rot = 45,
      top_annotation = col_ann_v1b,
      column_title = paste0(comp_name, ": Top 30 Up+Down DEGs by adj.P.Val (",
        nrow(top30_up), " up, ", nrow(top30_down), " down) - Comparison Libraries Only"),
      column_title_gp = gpar(fontsize = 12, fontface = "bold")
    )

    pdf_filename_v2b <- file.path(plotDir, paste0("heatmap_v2b_top30updown_adjpval_", comp_name, ".pdf"))
    pdf(pdf_filename_v2b, width = 12, height = max(8, nrow(top30_updown_v2) * 0.25))
    draw(ht_v2b, heatmap_legend_side = "right", annotation_legend_side = "right")
    dev.off()
  }

  # === VERSION 3: Top 30 -logFC and top 30 +logFC DEGs by abs(logFC), require adj.p.val < 0.05 ===
  # Get top 30 downregulated genes by absolute logFC (most negative, significant)
  top30_down_v3 <- sig_genes %>%
    filter(logFC < 0) %>%
    arrange(logFC) %>%  # Most negative first
    slice(1:min(30, nrow(.)))

  # Get top 30 upregulated genes by absolute logFC (most positive, significant)
  top30_up_v3 <- sig_genes %>%
    filter(logFC > 0) %>%
    arrange(desc(logFC)) %>%  # Most positive first
    slice(1:min(30, nrow(.)))

  # Combine up and down genes
  top30_updown_v3 <- rbind(top30_down_v3, top30_up_v3)

  if (nrow(top30_updown_v3) > 0) {
    # VERSION 3A: All libraries
    expr_data_v3a <- vwts.treatment$E[top30_updown_v3$gene, , drop = FALSE]
    expr_data_v3a_scaled <- t(scale(t(expr_data_v3a)))
    expr_data_v3a_ordered <- expr_data_v3a_scaled[, col_order_v1a, drop = FALSE]

    # Set dynamic color range based on data
    z_range_v3a <- range(expr_data_v3a_ordered, na.rm = TRUE)
    z_max_v3a <- max(abs(z_range_v3a))

    ht_v3a <- Heatmap(
      expr_data_v3a_ordered,
      name = "Z-score",
      col = colorRamp2(c(-z_max_v3a, 0, z_max_v3a), c("blue", "white", "red")),
      show_row_names = TRUE,
      row_names_gp = gpar(fontsize = 10),
      cluster_rows = TRUE,
      column_split = annotatedMetricsQC$stimulation[col_order_v1a],
      cluster_columns = TRUE,
      cluster_column_slices = FALSE,
      show_column_names = FALSE,
      column_names_gp = gpar(fontsize = 10),
      column_names_rot = 45,
      top_annotation = col_ann_v1a,
      column_title = paste0(comp_name, ": Top 30 Up+Down DEGs by |logFC| (",
        nrow(top30_up_v3), " up, ", nrow(top30_down_v3), " down) - All Libraries"),
      column_title_gp = gpar(fontsize = 12, fontface = "bold")
    )

    pdf_filename_v3a <- file.path(plotDir, paste0("heatmap_v3a_top30updown_abslogfc_", comp_name, ".pdf"))
    pdf(pdf_filename_v3a, width = 12, height = max(8, nrow(top30_updown_v3) * 0.25))
    draw(ht_v3a, heatmap_legend_side = "right", annotation_legend_side = "right")
    dev.off()

    # VERSION 3B: Only comparison-relevant libraries
    expr_data_v3b <- expr_data_v3a_scaled[, libs_subset, drop = FALSE]
    expr_data_v3b_scaled <- t(scale(t(expr_data_v3b)))
    expr_data_v3b_ordered <- expr_data_v3b_scaled[, col_order_v1b, drop = FALSE]

    # Set dynamic color range based on data
    z_range_v3b <- range(expr_data_v3b_ordered, na.rm = TRUE)
    z_max_v3b <- max(abs(z_range_v3b))

    ht_v3b <- Heatmap(
      expr_data_v3b_ordered,
      name = "Z-score",
      col = colorRamp2(c(-z_max_v3b, 0, z_max_v3b), c("blue", "white", "red")),
      show_row_names = TRUE,
      row_names_gp = gpar(fontsize = 10),
      cluster_rows = TRUE,
      column_split = anno_subset$stimulation[col_order_v1b],
      cluster_columns = TRUE,
      cluster_column_slices = FALSE,
      show_column_names = FALSE,
      column_names_gp = gpar(fontsize = 10),
      column_names_rot = 45,
      top_annotation = col_ann_v1b,
      column_title = paste0(comp_name, ": Top 30 Up+Down DEGs by |logFC| (",
        nrow(top30_up_v3), " up, ", nrow(top30_down_v3), " down) - Comparison Libraries Only"),
      column_title_gp = gpar(fontsize = 12, fontface = "bold")
    )

    pdf_filename_v3b <- file.path(plotDir, paste0("heatmap_v3b_top30updown_abslogfc_", comp_name, ".pdf"))
    pdf(pdf_filename_v3b, width = 12, height = max(8, nrow(top30_updown_v3) * 0.25))
    draw(ht_v3b, heatmap_legend_side = "right", annotation_legend_side = "right")
    dev.off()
  }

  cat("Completed heatmaps for:", comp_name, "\n\n")
}

cat("Completed all two-group DEG heatmaps\n")
```

```{r twoGroupDEGGSEAPlots}
# GSEA analysis for all two-group comparisons
library(clusterProfiler)
library(msigdbr)
library(enrichplot)

nTerms <- 15 # number of terms to include in the plots

# Function to create combined GSEA plots for any comparison
create_combined_gsea <- function(comparison_name, comparison_data) {
  cat("Running GSEA for:", comparison_name, "\n")

  # Get gene sets
  hs_sets <- msigdbr(species = "Homo sapiens", category = "H") %>%
    dplyr::select(gs_name, gene_symbol)

  GO_sets <- msigdbr(species = "Homo sapiens", category = "C5") %>%
    dplyr::select(gs_name, gene_symbol)

  # Prepare gene list - filter for genes with adj.P.Val <= 0.1 for GSEA input
  df_filtered <- comparison_data %>% dplyr::filter(adj.P.Val <= 0.05)

  if (nrow(df_filtered) == 0) {
    cat("No genes with adj.P.Val <= 0.1 found for", comparison_name, ". Skipping.\n")
    return(NULL)
  }

  geneList <- df_filtered$logFC
  names(geneList) <- df_filtered$gene
  geneList <- sort(geneList, decreasing = TRUE)

  cat("Gene list length for", comparison_name, ":", length(geneList), "\n")

  # Run GSEA for both gene sets
  tryCatch(
    {
      hallmark_gsea <- GSEA(geneList,
        TERM2GENE = hs_sets,
        minGSSize = 10,
        maxGSSize = 1000,
        pvalueCutoff = 0.05)

      go_gsea <- GSEA(geneList,
        TERM2GENE = GO_sets,
        minGSSize = 10,
        maxGSSize = 1000,
        pvalueCutoff = 0.05)

      # Select significant terms
      hallmark_significant <- hallmark_gsea@result %>%
        dplyr::filter(p.adjust < 0.05) %>%
        dplyr::arrange(p.adjust) %>%
        dplyr::slice(1:min(nTerms, nrow(.)))

      go_significant <- go_gsea@result %>%
        dplyr::filter(p.adjust < 0.05) %>%
        dplyr::arrange(p.adjust) %>%
        dplyr::slice(1:min(nTerms, nrow(.)))

      # Print debugging info
      cat("Hallmark significant terms (p.adjust < 0.05):", nrow(hallmark_significant), "\n")
      cat("GO significant terms (p.adjust < 0.05):", nrow(go_significant), "\n")

      # Check if we have any significant terms at all
      if (nrow(hallmark_significant) == 0 && nrow(go_significant) == 0) {
        cat("No significant terms found for", comparison_name, ". Creating empty plot.\n")
        empty_plot <- ggplot() +
          geom_text(aes(x = 0, y = 0, label = "No significant terms found"), size = 6) +
          theme_void() +
          ggtitle(paste0(comparison_name, ": No Significant Terms"))
        return(list(plot = empty_plot, data = data.frame(), comparison = comparison_name))
      }

      # Prepare data for plotting
      plot_data_list <- list()

      if (nrow(hallmark_significant) > 0) {
        hallmark_data <- hallmark_significant %>%
          dplyr::mutate(
            GeneSet = "Hallmark",
            Description_clean = paste0("H: ", gsub("HALLMARK_", "", Description))
          )
        plot_data_list[["Hallmark"]] <- hallmark_data
      }

      if (nrow(go_significant) > 0) {
        go_data <- go_significant %>%
          dplyr::mutate(
            GeneSet = "GO",
            Description_clean = paste0("GO: ", gsub("GOCC_|GOBP_|GOMF_", "", Description))
          )
        plot_data_list[["GO"]] <- go_data
      }

      # Only proceed if we have data to plot
      if (length(plot_data_list) == 0) {
        cat("No plot data for", comparison_name, "\n")
        return(NULL)
      }

      # Combine data - include setSize for dot size
      combined_data <- do.call(rbind, lapply(names(plot_data_list), function(gs) {
        plot_data_list[[gs]] %>% dplyr::select(ID, Description_clean, NES, p.adjust, setSize, GeneSet)
      }))

      # Order terms by NES magnitude within each gene set
      if ("Hallmark" %in% names(plot_data_list)) {
        hallmark_order <- plot_data_list[["Hallmark"]] %>%
          dplyr::arrange(desc(abs(NES))) %>%
          dplyr::pull(Description_clean)
      } else {
        hallmark_order <- character(0)
      }

      if ("GO" %in% names(plot_data_list)) {
        go_order <- plot_data_list[["GO"]] %>%
          dplyr::arrange(desc(abs(NES))) %>%
          dplyr::pull(Description_clean)
      } else {
        go_order <- character(0)
      }

      combined_data$Description_clean <- factor(combined_data$Description_clean,
        levels = c(hallmark_order, go_order))

      # Create the plot
      p <- ggplot(combined_data, aes(x = NES, y = Description_clean)) +
        geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
        geom_segment(aes(y = Description_clean, yend = Description_clean, x = 0, xend = NES),
          color = "black", linewidth = 0.5) +
        geom_point(aes(
          size = setSize,
          fill = -log10(p.adjust)
        ),
        shape = 21,
        color = "black",
        stroke = 0.5) +
        facet_wrap(~GeneSet, scales = "free_y", nrow = 1) +
        scale_fill_gradient2(
          low = "lightblue",
          mid = "white",
          high = "red",
          midpoint = -log10(0.05),  # White at significance threshold
          name = "-log10(Adj.P)",
          guide = guide_colorbar(title.position = "top")
        ) +
        scale_size_continuous(
          name = "Gene Set Size",
          range = c(2, 8)
        ) +
        labs(
          x = "Normalized Enrichment Score (NES)",
          y = "",
          title = paste0(comparison_name, ": Hallmark and GO Gene Sets")
        ) +
        theme_bw() +
        theme(
          strip.background = element_rect(fill = "white", color = "black", linewidth = 1),
          strip.text = element_text(face = "bold", size = 14),
          plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
          axis.text.y = element_text(size = 10),
          axis.text.x = element_text(size = 12),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          legend.position = "right",
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_line(color = "gray95"),
          panel.grid.major.y = element_blank(),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
        )

      # Add facetted scales if we have both gene sets
      if (length(unique(combined_data$GeneSet)) > 1) {
        p <- p + ggh4x::facetted_pos_scales(
          y = list(
            scale_y_discrete(position = "left"),   # First facet on left
            scale_y_discrete(position = "right")   # Second facet on right
          )
        )
      }

      return(list(plot = p, data = combined_data, comparison = comparison_name,
        hallmark_gsea = hallmark_gsea, go_gsea = go_gsea))

    },
    error = function(e) {
      cat("Error in GSEA for", comparison_name, ":", e$message, "\n")
      return(NULL)
    })
}

# Run GSEA for all two-group comparisons
gsea_results <- list()

for (comp_name in names(comparisons)) {
  cat("\n=== Processing", comp_name, "===\n")

  result <- create_combined_gsea(comp_name, comparisons[[comp_name]])

  if (!is.null(result)) {
    gsea_results[[comp_name]] <- result

    # Save the plot
    pdf_filename <- file.path(plotDir, paste0("GSEA_", comp_name, ".pdf"))
    png_filename <- file.path(plotDir, paste0("GSEA_", comp_name, ".png"))

    # Save PDF
    pdf(pdf_filename, width = 18, height = 8)
    print(result$plot)
    dev.off()

    # Save PNG
    png(png_filename, width = 18, height = 8, units = "in", res = 600)
    print(result$plot)
    dev.off()

    cat("Saved plots for", comp_name, "\n")

    # Save GSEA results as CSV
    if (!is.null(result$hallmark_gsea) && nrow(result$hallmark_gsea@result) > 0) {
      write.csv(result$hallmark_gsea@result,
        file = file.path(outputDataDir, paste0("GSEA_Hallmark_", comp_name, ".csv")),
        row.names = FALSE)
    }

    if (!is.null(result$go_gsea) && nrow(result$go_gsea@result) > 0) {
      write.csv(result$go_gsea@result,
        file = file.path(outputDataDir, paste0("GSEA_GO_", comp_name, ".csv")),
        row.names = FALSE)
    }
  }
}

cat("\n=== GSEA Analysis Complete ===\n")
cat("Total comparisons processed:", length(gsea_results), "\n")
cat("Results saved to:", plotDir, "\n")
```


# are there responses to these stimuli?
# IL4 - NoTx
# IL4_IL17 - NoTx
# TNFa - NoTx
# TNFa_IL17 - NoTx

# Do responses to stimuli change with 3G9 (anti-avb6) or anti-TGFb?
# (IL4 - NoTx) vs (IL4_3G9 - NoTx)
# (IL4 - NoTx) vs (IL4_antiTGFb - NoTx)
# (IL4_IL17 - NoTx) vs (IL_IL17_3G9 - NoTx)
# (IL4_IL17 - NoTx) vs (IL_IL17_antiTGFb - NoTx)
# (TNFa - NoTx) vs (TNFa_3G9 vs NoTx)
# (TNFa - NoTx) vs (TNFa_antiTGFb - NoTx)
# (TNFa_IL17 - NoTx) vs (TNFa_IL17_3G9 - NoTx)
# (TNFa_IL17 - NoTx) vs (TNFa_IL17_antiTGFb - NoTx)

# Are there any DEGs shared between 3G9 and anti-TGFb?
# (IL4_3G9 - NoTx) vs (IL4_antiTGFb - NoTx)
# (IL_IL17_3G9 - NoTx) vs (IL_IL17_antiTGFb - NoTx)
# (TNFa_3G9 vs NoTx) vs (TNFa_antiTGFb - NoTx)
# (TNFa_IL17_3G9 - NoTx) vs (TNFa_IL17_antiTGFb - NoTx)
#  scatterplot for each of these comparisons
#  Heatmaps of shared DEGs--separate for stim condition




# After we do 2-group vs 2-group analyses, we should really try a 4-way interaction term explicitly modeled

```{r logFCScatterplotsVennDiagrams3G9vsantiTGFbEffect}
# Define comparison pairs to assess how stimulation signatures change with 3G9 or antiTGFb
comparison_pairs <- list(
  list(base = "IL4_vs_NoTx", treat1 = "IL4_3G9_vs_NoTx", treat2 = "IL4_antiTGFb_vs_NoTx",
    stim = "IL4", label1 = "3G9", label2 = "antiTGFb"),
  list(base = "IL4_IL17_vs_NoTx", treat1 = "IL4_IL17_3G9_vs_NoTx", treat2 = "IL4_IL17_antiTGFb_vs_NoTx",
    stim = "IL4_IL17", label1 = "3G9", label2 = "antiTGFb"),
  list(base = "TNFa_vs_NoTx", treat1 = "TNFa_3G9_vs_NoTx", treat2 = "TNFa_antiTGFb_vs_NoTx",
    stim = "TNFa", label1 = "3G9", label2 = "antiTGFb"),
  list(base = "TNFa_IL17_vs_NoTx", treat1 = "TNFa_IL17_3G9_vs_NoTx", treat2 = "TNFa_IL17_antiTGFb_vs_NoTx",
    stim = "TNFa_IL17", label1 = "3G9", label2 = "antiTGFb")
)

# Create scatterplots and Venn diagrams for each comparison pair
for (pair in comparison_pairs) {
  # Extract comparison data
  base_comp <- comparisons[[pair$base]]
  treat1_comp <- comparisons[[pair$treat1]]
  treat2_comp <- comparisons[[pair$treat2]]

  # === SCATTERPLOT 1: treat1 vs treat2 ===
  # Merge the two treatment comparisons
  merged_df <- merge(treat1_comp, treat2_comp, by = "gene", suffixes = c(paste0("_", pair$label1), paste0("_", pair$label2)))

  # Flag genes that are significant in either contrast
  merged_df <- merged_df %>%
    mutate(
      sig = (get(paste0("adj.P.Val_", pair$label1)) <= 0.05) | (get(paste0("adj.P.Val_", pair$label2)) <= 0.05),
      sig_both = (get(paste0("adj.P.Val_", pair$label1)) <= 0.05) & (get(paste0("adj.P.Val_", pair$label2)) <= 0.05)
    ) %>%
    # Calculate distance from y=x diagonal for discordance ranking
    mutate(
      distance_from_diagonal = abs(get(paste0("logFC_", pair$label2)) - get(paste0("logFC_", pair$label1)))
    )

  # Get top 30 most discordant genes among those significant in either comparison
  top30_discordant <- merged_df %>%
    filter(sig) %>%
    arrange(desc(distance_from_diagonal)) %>%
    slice(1:30)

  # Create scatterplot
  p_scatter <- ggplot(merged_df, aes(x = get(paste0("logFC_", pair$label1)), y = get(paste0("logFC_", pair$label2)))) +
    # Non-significant points
    geom_point(data = filter(merged_df, !sig),
      aes(fill = get(paste0("adj.P.Val_", pair$label2)), color = get(paste0("adj.P.Val_", pair$label1))),
      shape = 21, size = 3, stroke = 1, alpha = 0.5) +
    # Significant points
    geom_point(data = filter(merged_df, sig),
      aes(fill = get(paste0("adj.P.Val_", pair$label2)), color = get(paste0("adj.P.Val_", pair$label1))),
      shape = 21, size = 3, stroke = 1) +
    # Label only top 30 most discordant genes significant in either comparison
    geom_text_repel(data = top30_discordant,
      aes(label = gene),
      size = 3, max.overlaps = Inf) +
    scale_fill_gradient(low = "blue", high = "grey", limits = c(0, 0.05), oob = scales::squish,
      name = paste0("adj.P.Val (", pair$label2, ")")) +
    scale_color_gradient(low = "red", high = "grey", limits = c(0, 0.05), oob = scales::squish,
      name = paste0("adj.P.Val (", pair$label1, ")")) +
    coord_equal() +
    geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
    geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
    labs(x = paste0("logFC: ", pair$stim, "_", pair$label1, " - NoTx"),
      y = paste0("logFC: ", pair$stim, "_", pair$label2, " - NoTx"),
      title = paste0("Scatterplot: ", pair$stim, " response with ", pair$label1, " vs ", pair$label2)) +
    theme_bw() + theme(text = element_text(size = 12))

  # Save scatterplot
  pdf_filename <- file.path(plotDir, paste0("logFC_scatterplot_", pair$stim, "_", pair$label1, "_vs_", pair$label2, ".pdf"))
  pdf(file = pdf_filename, width = 12, height = 8)
  print(p_scatter)
  dev.off()

  # === VENN DIAGRAM ===
  # Get significant genes for each comparison
  sig_treat1 <- treat1_comp %>% filter(adj.P.Val <= 0.05) %>% pull(gene)
  sig_treat2 <- treat2_comp %>% filter(adj.P.Val <= 0.05) %>% pull(gene)

  # Create Venn diagram
  venn_list <- list()
  venn_list[[paste0(pair$stim, "_", pair$label1)]] <- sig_treat1
  venn_list[[paste0(pair$stim, "_", pair$label2)]] <- sig_treat2

  # Plot Venn diagram
  p_venn <- ggvenn(venn_list,
    fill_color = c("lightblue", "lightcoral"),
    stroke_size = 0.5, set_name_size = 4)

  # Save Venn diagram
  pdf_filename_venn <- file.path(plotDir, paste0("venn_diagram_", pair$stim, "_", pair$label1, "_vs_", pair$label2, ".pdf"))
  pdf(file = pdf_filename_venn, width = 8, height = 6)
  print(p_venn)
  dev.off()

  # === SAVE VENN CATEGORIES TO XLSX ===
  # Get intersection and unique genes
  intersection <- intersect(sig_treat1, sig_treat2)
  unique_treat1 <- setdiff(sig_treat1, sig_treat2)
  unique_treat2 <- setdiff(sig_treat2, sig_treat1)

  # Create workbook
  wb <- createWorkbook()

  # Add worksheets with gene lists
  addWorksheet(wb, "Intersection")
  addWorksheet(wb, paste0("Unique_", pair$label1))
  addWorksheet(wb, paste0("Unique_", pair$label2))
  addWorksheet(wb, "All_sig_genes")

  # Write data
  if (length(intersection) > 0) {
    intersection_df <- treat1_comp %>% filter(gene %in% intersection) %>%
      select(gene, logFC, adj.P.Val) %>%
      rename_with(~ paste0(.x, "_", pair$label1), c(logFC, adj.P.Val)) %>%
      left_join(treat2_comp %>% filter(gene %in% intersection) %>%
        select(gene, logFC, adj.P.Val) %>%
        rename_with(~ paste0(.x, "_", pair$label2), c(logFC, adj.P.Val)), by = "gene")
    writeData(wb, "Intersection", intersection_df)
  }

  if (length(unique_treat1) > 0) {
    unique_treat1_df <- treat1_comp %>% filter(gene %in% unique_treat1) %>%
      select(gene, logFC, adj.P.Val)
    writeData(wb, paste0("Unique_", pair$label1), unique_treat1_df)
  }

  if (length(unique_treat2) > 0) {
    unique_treat2_df <- treat2_comp %>% filter(gene %in% unique_treat2) %>%
      select(gene, logFC, adj.P.Val)
    writeData(wb, paste0("Unique_", pair$label2), unique_treat2_df)
  }

  # All significant genes summary
  all_sig_summary <- data.frame(
    Category = c(paste0("Total_sig_genes_", pair$label1), paste0("Total_sig_genes_", pair$label2),
      "Intersection", paste0("Unique_", pair$label1), paste0("Unique_", pair$label2)),
    Count = c(length(sig_treat1), length(sig_treat2), length(intersection),
      length(unique_treat1), length(unique_treat2))
  )
  writeData(wb, "All_sig_genes", all_sig_summary)

  # Save workbook
  xlsx_filename <- file.path(outputDataDir, paste0("venn_categories_", pair$stim, "_", pair$label1, "_vs_", pair$label2, ".xlsx"))
  saveWorkbook(wb, xlsx_filename, overwrite = TRUE)

  cat("Completed analysis for:", pair$stim, "comparing", pair$label1, "vs", pair$label2, "\n")
}
```

```{r heatmaps3G9vsantiTGFbEffect}
# Generate heatmaps for shared DEGs between 3G9 and antiTGFb effects
# Load the 4 Excel files containing shared DEGs and create heatmaps

library(ComplexHeatmap)
library(circlize)
library(openxlsx)

# Define the 4 comparison pairs that have shared DEGs
shared_pairs <- list(
  list(stim = "IL4", label1 = "3G9", label2 = "antiTGFb"),
  list(stim = "IL4_IL17", label1 = "3G9", label2 = "antiTGFb"),
  list(stim = "TNFa", label1 = "3G9", label2 = "antiTGFb"),
  list(stim = "TNFa_IL17", label1 = "3G9", label2 = "antiTGFb")
)

# Create heatmaps for each shared DEG set
for (pair in shared_pairs) {
  cat("Creating heatmaps for shared DEGs:", pair$stim, "comparing", pair$label1, "vs", pair$label2, "\n")

  # Read the Excel file with shared DEGs
  xlsx_filename <- file.path(outputDataDir, paste0("venn_categories_", pair$stim, "_", pair$label1, "_vs_", pair$label2, ".xlsx"))

  if (!file.exists(xlsx_filename)) {
    cat("Warning: Excel file not found:", xlsx_filename, "\n")
    next
  }

  # Read the intersection genes sheet
  shared_genes <- read.xlsx(xlsx_filename, sheet = "Intersection")

  if (nrow(shared_genes) == 0) {
    cat("No shared genes found for:", pair$stim, "\n")
    next
  }

  cat("Found", nrow(shared_genes), "shared genes for", pair$stim, "\n")

  # Get expression data for these genes
  gene_names <- shared_genes$gene
  expr_data <- vwts.treatment$E[gene_names, , drop = FALSE]

  # Z-score scale the expression data by gene (row-wise)
  expr_data_scaled <- t(scale(t(expr_data)))

  # Define relevant treatment groups for this comparison
  # For each stimulation, include: base stim, stim_3G9, stim_antiTGFb, and NoTx
  relevant_treatments <- c(pair$stim, paste0(pair$stim, "_3G9"), paste0(pair$stim, "_antiTGFb"), "NoTx")

  # Get libraries for version b (only those involved in this specific comparison)
  libs_subset <- annotatedMetricsQC$libid[annotatedMetricsQC$treatment %in% relevant_treatments]
  anno_subset <- annotatedMetricsQC[annotatedMetricsQC$libid %in% libs_subset, ]

  # === VERSION A: All libraries ===

  # Create column annotation based on treatment and stimulation
  col_ann_a <- HeatmapAnnotation(
    Treatment = annotatedMetricsQC$treatment,
    Stimulation = annotatedMetricsQC$stimulation,
    col = list(
      Treatment = palTreatment,
      Stimulation = palStimulation
    ),
    annotation_name_gp = gpar(fontsize = 10),
    annotation_legend_param = list(
      Treatment = list(title_gp = gpar(fontsize = 12), labels_gp = gpar(fontsize = 10)),
      Stimulation = list(title_gp = gpar(fontsize = 12), labels_gp = gpar(fontsize = 10))
    )
  )

  # Order columns by stimulation, then by treatment within stimulation
  col_order_a <- order(annotatedMetricsQC$stimulation, annotatedMetricsQC$treatment)
  expr_data_ordered_a <- expr_data_scaled[, col_order_a, drop = FALSE]

  # Set dynamic color range
  z_range_a <- range(expr_data_ordered_a, na.rm = TRUE)
  z_max_a <- max(abs(z_range_a))

  # Create the heatmap version A
  ht_a <- Heatmap(
    expr_data_ordered_a,
    name = "Z-score",
    col = colorRamp2(c(-z_max_a, 0, z_max_a), c("blue", "white", "red")),

    # Row settings - cluster genes but don't show gene names
    show_row_names = FALSE,
    cluster_rows = TRUE,
    row_dend_width = unit(10, "mm"),

    # Column settings - group by stimulation, cluster within groups
    column_split = annotatedMetricsQC$stimulation[col_order_a],
    cluster_columns = TRUE,
    cluster_column_slices = FALSE,
    column_gap = unit(2, "mm"),
    show_column_names = FALSE,
    column_names_gp = gpar(fontsize = 10),
    column_names_rot = 45,

    # Add column annotation
    top_annotation = col_ann_a,

    # Heatmap body settings
    heatmap_legend_param = list(
      title = "Z-score",
      title_gp = gpar(fontsize = 12),
      labels_gp = gpar(fontsize = 10),
      legend_height = unit(4, "cm")
    ),

    # Overall settings
    row_title = paste0("Shared DEGs (n=", nrow(shared_genes), ")"),
    row_title_gp = gpar(fontsize = 12),
    column_title = paste0(pair$stim, ": Shared DEGs between ", pair$label1, " and ", pair$label2, " effects (All Libraries)"),
    column_title_gp = gpar(fontsize = 14, fontface = "bold")
  )

  # Save the heatmap version A as PDF
  pdf_filename_a <- file.path(plotDir, paste0("heatmap_a_shared_DEGs_", pair$stim, "_", pair$label1, "_vs_", pair$label2, ".pdf"))

  pdf(pdf_filename_a, width = 12, height = 8)
  draw(ht_a, heatmap_legend_side = "right", annotation_legend_side = "right")
  dev.off()

  cat("Saved heatmap A:", pdf_filename_a, "\n")

  # === VERSION B: Only comparison-relevant libraries ===

  # Subset expression data to relevant libraries and re-scale
  expr_data_subset <- expr_data_scaled[, libs_subset, drop = FALSE]
  expr_data_subset_scaled <- t(scale(t(expr_data_subset)))  # Re-scale for subset

  # Create column annotation for subset
  col_ann_b <- HeatmapAnnotation(
    Treatment = anno_subset$treatment,
    Stimulation = anno_subset$stimulation,
    col = list(
      Treatment = palTreatment,
      Stimulation = palStimulation
    ),
    annotation_name_gp = gpar(fontsize = 10),
    annotation_legend_param = list(
      Treatment = list(title_gp = gpar(fontsize = 12), labels_gp = gpar(fontsize = 10)),
      Stimulation = list(title_gp = gpar(fontsize = 12), labels_gp = gpar(fontsize = 10))
    )
  )

  # Order columns by stimulation, then by treatment within stimulation
  col_order_b <- order(anno_subset$stimulation, anno_subset$treatment)
  expr_data_ordered_b <- expr_data_subset_scaled[, col_order_b, drop = FALSE]

  # Set dynamic color range
  z_range_b <- range(expr_data_ordered_b, na.rm = TRUE)
  z_max_b <- max(abs(z_range_b))

  # Create the heatmap version B
  ht_b <- Heatmap(
    expr_data_ordered_b,
    name = "Z-score",
    col = colorRamp2(c(-z_max_b, 0, z_max_b), c("blue", "white", "red")),

    # Row settings - cluster genes but don't show gene names
    show_row_names = FALSE,
    cluster_rows = TRUE,
    row_dend_width = unit(10, "mm"),

    # Column settings - group by stimulation, cluster within groups
    column_split = anno_subset$stimulation[col_order_b],
    cluster_columns = TRUE,
    cluster_column_slices = FALSE,
    column_gap = unit(2, "mm"),
    show_column_names = FALSE,
    column_names_gp = gpar(fontsize = 10),
    column_names_rot = 45,

    # Add column annotation
    top_annotation = col_ann_b,

    # Heatmap body settings
    heatmap_legend_param = list(
      title = "Z-score",
      title_gp = gpar(fontsize = 12),
      labels_gp = gpar(fontsize = 10),
      legend_height = unit(4, "cm")
    ),

    # Overall settings
    row_title = paste0("Shared DEGs (n=", nrow(shared_genes), ")"),
    row_title_gp = gpar(fontsize = 12),
    column_title = paste0(pair$stim, ": Shared DEGs between ", pair$label1, " and ", pair$label2, " effects (Comparison Libraries Only)"),
    column_title_gp = gpar(fontsize = 14, fontface = "bold")
  )

  # Save the heatmap version B as PDF
  pdf_filename_b <- file.path(plotDir, paste0("heatmap_b_shared_DEGs_", pair$stim, "_", pair$label1, "_vs_", pair$label2, ".pdf"))

  pdf(pdf_filename_b, width = 10, height = 8)
  draw(ht_b, heatmap_legend_side = "right", annotation_legend_side = "right")
  dev.off()

  cat("Saved heatmap B:", pdf_filename_b, "\n\n")
}

cat("Completed all shared DEG heatmaps (A and B versions)\n")

# ============================version with just top 50 genes============================================================

# Create second version of heatmaps: Top 50 genes by adj.P.Val with gene names labeled
cat("Creating top 50 gene heatmaps with gene labels (A and B versions)...\n")

for (pair in shared_pairs) {
  cat("Creating top 50 gene heatmaps for:", pair$stim, "comparing", pair$label1, "vs", pair$label2, "\n")

  # Read the Excel file with shared DEGs
  xlsx_filename <- file.path(outputDataDir, paste0("venn_categories_", pair$stim, "_", pair$label1, "_vs_", pair$label2, ".xlsx"))

  if (!file.exists(xlsx_filename)) {
    cat("Warning: Excel file not found:", xlsx_filename, "\n")
    next
  }

  # Read the intersection genes sheet
  shared_genes <- read.xlsx(xlsx_filename, sheet = "Intersection")

  if (nrow(shared_genes) == 0) {
    cat("No shared genes found for:", pair$stim, "\n")
    next
  }

  # Sort by the minimum adjusted p-value from either comparison and take top 50
  shared_genes$min_adj_pval <- pmin(
    shared_genes[[paste0("adj.P.Val_", pair$label1)]],
    shared_genes[[paste0("adj.P.Val_", pair$label2)]]
  )
  shared_genes_top50 <- shared_genes %>%
    arrange(min_adj_pval) %>%
    slice(1:min(50, nrow(shared_genes)))

  cat("Selected top", nrow(shared_genes_top50), "genes for", pair$stim, "\n")

  # Get expression data for these top genes
  gene_names_top50 <- shared_genes_top50$gene
  expr_data_top50 <- vwts.treatment$E[gene_names_top50, , drop = FALSE]

  # Z-score scale the expression data by gene (row-wise)
  expr_data_top50_scaled <- t(scale(t(expr_data_top50)))

  # Define relevant treatment groups for this comparison
  relevant_treatments <- c(pair$stim, paste0(pair$stim, "_3G9"), paste0(pair$stim, "_antiTGFb"), "NoTx")

  # Get libraries for version b (only those involved in this specific comparison)
  libs_subset <- annotatedMetricsQC$libid[annotatedMetricsQC$treatment %in% relevant_treatments]
  anno_subset <- annotatedMetricsQC[annotatedMetricsQC$libid %in% libs_subset, ]

  # === VERSION A: All libraries (Top 50 with gene names) ===

  # Create column annotation based on treatment and stimulation
  col_ann_top50_a <- HeatmapAnnotation(
    Treatment = annotatedMetricsQC$treatment,
    Stimulation = annotatedMetricsQC$stimulation,
    col = list(
      Treatment = palTreatment,
      Stimulation = palStimulation
    ),
    annotation_name_gp = gpar(fontsize = 10),
    annotation_legend_param = list(
      Treatment = list(title_gp = gpar(fontsize = 12), labels_gp = gpar(fontsize = 10)),
      Stimulation = list(title_gp = gpar(fontsize = 12), labels_gp = gpar(fontsize = 10))
    )
  )

  # Order columns by stimulation, then by treatment within stimulation
  col_order_top50_a <- order(annotatedMetricsQC$stimulation, annotatedMetricsQC$treatment)
  expr_data_top50_ordered_a <- expr_data_top50_scaled[, col_order_top50_a, drop = FALSE]

  # Set dynamic color range
  z_range_top50_a <- range(expr_data_top50_ordered_a, na.rm = TRUE)
  z_max_top50_a <- max(abs(z_range_top50_a))

  # Create the heatmap with gene names labeled - Version A
  ht_top50_a <- Heatmap(
    expr_data_top50_ordered_a,
    name = "Z-score",
    col = colorRamp2(c(-z_max_top50_a, 0, z_max_top50_a), c("blue", "white", "red")),

    # Row settings - cluster genes and show gene names
    show_row_names = TRUE,
    row_names_gp = gpar(fontsize = 10),
    cluster_rows = TRUE,
    row_dend_width = unit(10, "mm"),

    # Column settings - group by stimulation, cluster within groups
    column_split = annotatedMetricsQC$stimulation[col_order_top50_a],
    cluster_columns = TRUE,
    cluster_column_slices = FALSE,
    column_gap = unit(2, "mm"),
    show_column_names = FALSE,
    column_names_gp = gpar(fontsize = 10),
    column_names_rot = 45,

    # Add column annotation
    top_annotation = col_ann_top50_a,

    # Heatmap body settings
    heatmap_legend_param = list(
      title = "Z-score",
      title_gp = gpar(fontsize = 12),
      labels_gp = gpar(fontsize = 10),
      legend_height = unit(4, "cm")
    ),

    # Overall settings
    row_title = paste0("Top ", nrow(shared_genes_top50), " Shared DEGs (by adj.P.Val)"),
    row_title_gp = gpar(fontsize = 12),
    column_title = paste0(pair$stim, ": Top Shared DEGs between ", pair$label1, " and ", pair$label2, " effects (All Libraries)"),
    column_title_gp = gpar(fontsize = 14, fontface = "bold")
  )

  # Save the top 50 heatmap version A as PDF
  pdf_filename_top50_a <- file.path(plotDir, paste0("heatmap_a_top50_shared_DEGs_", pair$stim, "_", pair$label1, "_vs_", pair$label2, ".pdf"))

  pdf(pdf_filename_top50_a, width = 14, height = 10)
  draw(ht_top50_a, heatmap_legend_side = "right", annotation_legend_side = "right")
  dev.off()

  cat("Saved top 50 heatmap A:", pdf_filename_top50_a, "\n")

  # === VERSION B: Only comparison-relevant libraries (Top 50 with gene names) ===

  # Subset expression data to relevant libraries and re-scale
  expr_data_top50_subset <- expr_data_top50_scaled[, libs_subset, drop = FALSE]
  expr_data_top50_subset_scaled <- t(scale(t(expr_data_top50_subset)))  # Re-scale for subset

  # Create column annotation for subset
  col_ann_top50_b <- HeatmapAnnotation(
    Treatment = anno_subset$treatment,
    Stimulation = anno_subset$stimulation,
    col = list(
      Treatment = palTreatment,
      Stimulation = palStimulation
    ),
    annotation_name_gp = gpar(fontsize = 10),
    annotation_legend_param = list(
      Treatment = list(title_gp = gpar(fontsize = 12), labels_gp = gpar(fontsize = 10)),
      Stimulation = list(title_gp = gpar(fontsize = 12), labels_gp = gpar(fontsize = 10))
    )
  )

  # Order columns by stimulation, then by treatment within stimulation
  col_order_top50_b <- order(anno_subset$stimulation, anno_subset$treatment)
  expr_data_top50_ordered_b <- expr_data_top50_subset_scaled[, col_order_top50_b, drop = FALSE]

  # Set dynamic color range
  z_range_top50_b <- range(expr_data_top50_ordered_b, na.rm = TRUE)
  z_max_top50_b <- max(abs(z_range_top50_b))

  # Create the heatmap with gene names labeled - Version B
  ht_top50_b <- Heatmap(
    expr_data_top50_ordered_b,
    name = "Z-score",
    col = colorRamp2(c(-z_max_top50_b, 0, z_max_top50_b), c("blue", "white", "red")),

    # Row settings - cluster genes and show gene names
    show_row_names = TRUE,
    row_names_gp = gpar(fontsize = 10),
    cluster_rows = TRUE,
    row_dend_width = unit(10, "mm"),

    # Column settings - group by stimulation, cluster within groups
    column_split = anno_subset$stimulation[col_order_top50_b],
    cluster_columns = TRUE,
    cluster_column_slices = FALSE,
    column_gap = unit(2, "mm"),
    show_column_names = FALSE,
    column_names_gp = gpar(fontsize = 10),
    column_names_rot = 45,

    # Add column annotation
    top_annotation = col_ann_top50_b,

    # Heatmap body settings
    heatmap_legend_param = list(
      title = "Z-score",
      title_gp = gpar(fontsize = 12),
      labels_gp = gpar(fontsize = 10),
      legend_height = unit(4, "cm")
    ),

    # Overall settings
    row_title = paste0("Top ", nrow(shared_genes_top50), " Shared DEGs (by adj.P.Val)"),
    row_title_gp = gpar(fontsize = 12),
    column_title = paste0(pair$stim, ": Top Shared DEGs between ", pair$label1, " and ", pair$label2, " effects (Comparison Libraries Only)"),
    column_title_gp = gpar(fontsize = 14, fontface = "bold")
  )

  # Save the top 50 heatmap version B as PDF
  pdf_filename_top50_b <- file.path(plotDir, paste0("heatmap_b_top50_shared_DEGs_", pair$stim, "_", pair$label1, "_vs_", pair$label2, ".pdf"))

  pdf(pdf_filename_top50_b, width = 12, height = 10)
  draw(ht_top50_b, heatmap_legend_side = "right", annotation_legend_side = "right")
  dev.off()

  cat("Saved top 50 heatmap B:", pdf_filename_top50_b, "\n\n")
}

cat("Completed all top 50 shared DEG heatmaps with gene labels (A and B versions)\n")
```

```{r stimulationSpecificSignaturesAlteredBy3G9_OR_antiTGFb}
# Define the 8 specific comparison pairs as requested
# These compare how modifiers (3G9 or antiTGFb) affect each stimulation condition
modifier_effect_pairs <- list(
  # (IL4 - NoTx) vs (IL4_3G9 - NoTx)
  list(comp1 = "IL4_vs_NoTx", comp2 = "IL4_3G9_vs_NoTx",
    label1 = "IL4", label2 = "IL4_3G9"),
  # (IL4 - NoTx) vs (IL4_antiTGFb - NoTx)
  list(comp1 = "IL4_vs_NoTx", comp2 = "IL4_antiTGFb_vs_NoTx",
    label1 = "IL4", label2 = "IL4_antiTGFb"),
  # (IL4_IL17 - NoTx) vs (IL4_IL17_3G9 - NoTx)
  list(comp1 = "IL4_IL17_vs_NoTx", comp2 = "IL4_IL17_3G9_vs_NoTx",
    label1 = "IL4_IL17", label2 = "IL4_IL17_3G9"),
  # (IL4_IL17 - NoTx) vs (IL4_IL17_antiTGFb - NoTx)
  list(comp1 = "IL4_IL17_vs_NoTx", comp2 = "IL4_IL17_antiTGFb_vs_NoTx",
    label1 = "IL4_IL17", label2 = "IL4_IL17_antiTGFb"),
  # (TNFa - NoTx) vs (TNFa_3G9 - NoTx)
  list(comp1 = "TNFa_vs_NoTx", comp2 = "TNFa_3G9_vs_NoTx",
    label1 = "TNFa", label2 = "TNFa_3G9"),
  # (TNFa - NoTx) vs (TNFa_antiTGFb - NoTx)
  list(comp1 = "TNFa_vs_NoTx", comp2 = "TNFa_antiTGFb_vs_NoTx",
    label1 = "TNFa", label2 = "TNFa_antiTGFb"),
  # (TNFa_IL17 - NoTx) vs (TNFa_IL17_3G9 - NoTx)
  list(comp1 = "TNFa_IL17_vs_NoTx", comp2 = "TNFa_IL17_3G9_vs_NoTx",
    label1 = "TNFa_IL17", label2 = "TNFa_IL17_3G9"),
  # (TNFa_IL17 - NoTx) vs (TNFa_IL17_antiTGFb - NoTx)
  list(comp1 = "TNFa_IL17_vs_NoTx", comp2 = "TNFa_IL17_antiTGFb_vs_NoTx",
    label1 = "TNFa_IL17", label2 = "TNFa_IL17_antiTGFb")
)

# Create scatterplots and Venn diagrams for each modifier effect comparison
for (pair in modifier_effect_pairs) {
  # Extract comparison data
  comp1_data <- comparisons[[pair$comp1]]
  comp2_data <- comparisons[[pair$comp2]]

  # === SCATTERPLOT ===
  # Merge the two comparisons
  merged_df <- merge(comp1_data, comp2_data, by = "gene", suffixes = c("_comp1", "_comp2"))

  # Flag genes that are significant in either contrast
  merged_df <- merged_df %>%
    mutate(
      sig = (adj.P.Val_comp1 <= 0.05) | (adj.P.Val_comp2 <= 0.05),
      sig_both = (adj.P.Val_comp1 <= 0.05) & (adj.P.Val_comp2 <= 0.05)
    ) %>%
    # Calculate distance from y=x diagonal for discordance ranking
    mutate(
      distance_from_diagonal = abs(logFC_comp2 - logFC_comp1)
    )

  # Get top 30 most discordant genes among those significant in either comparison
  top30_discordant <- merged_df %>%
    filter(sig) %>%
    arrange(desc(distance_from_diagonal)) %>%
    slice(1:30)

  # Create scatterplot
  p_scatter <- ggplot(merged_df, aes(x = logFC_comp1, y = logFC_comp2)) +
    # Non-significant points
    geom_point(data = filter(merged_df, !sig),
      aes(fill = adj.P.Val_comp2, color = adj.P.Val_comp1),
      shape = 21, size = 3, stroke = 1, alpha = 0.5) +
    # Significant points
    geom_point(data = filter(merged_df, sig),
      aes(fill = adj.P.Val_comp2, color = adj.P.Val_comp1),
      shape = 21, size = 3, stroke = 1) +
    # Label only top 30 most discordant genes significant in either comparison
    geom_text_repel(data = top30_discordant,
      aes(label = gene),
      size = 3, max.overlaps = Inf) +
    scale_fill_gradient(low = "blue", high = "grey", limits = c(0, 0.05), oob = scales::squish,
      name = paste0("adj.P.Val (", pair$label2, ")")) +
    scale_color_gradient(low = "red", high = "grey", limits = c(0, 0.05), oob = scales::squish,
      name = paste0("adj.P.Val (", pair$label1, ")")) +
    coord_equal() +
    geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
    geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
    labs(x = paste0("logFC: ", pair$label1, " - NoTx"),
      y = paste0("logFC: ", pair$label2, " - NoTx"),
      title = paste0(pair$label1, " vs ", pair$label2, " response")) +
    theme_bw() + theme(text = element_text(size = 12))

  # Save scatterplot
  pdf_filename <- file.path(plotDir, paste0("logFC_scatterplot_", pair$label1, "_vs_", pair$label2, ".pdf"))
  pdf(file = pdf_filename, width = 12, height = 8)
  print(p_scatter)
  dev.off()

  # === VENN DIAGRAM ===
  # Get significant genes for each comparison
  sig_comp1 <- comp1_data %>% filter(adj.P.Val <= 0.05) %>% pull(gene)
  sig_comp2 <- comp2_data %>% filter(adj.P.Val <= 0.05) %>% pull(gene)

  # Create Venn diagram
  venn_list <- list()
  venn_list[[pair$label1]] <- sig_comp1
  venn_list[[pair$label2]] <- sig_comp2

  # Plot Venn diagram
  p_venn <- ggvenn(venn_list,
    fill_color = c("lightblue", "lightpink"),
    stroke_size = 0.5, set_name_size = 4)

  # Save Venn diagram
  pdf_filename_venn <- file.path(plotDir, paste0("venn_diagram_", pair$label1, "_vs_", pair$label2, ".pdf"))
  pdf(file = pdf_filename_venn, width = 8, height = 6)
  print(p_venn)
  dev.off()

  # === SAVE VENN CATEGORIES TO XLSX ===
  # Get intersection and unique genes
  intersection <- intersect(sig_comp1, sig_comp2)
  unique_comp1 <- setdiff(sig_comp1, sig_comp2)
  unique_comp2 <- setdiff(sig_comp2, sig_comp1)

  # Create workbook
  wb <- createWorkbook()

  # Add worksheets with gene lists
  addWorksheet(wb, "Intersection")
  addWorksheet(wb, paste0("Unique_", pair$label1))
  addWorksheet(wb, paste0("Unique_", pair$label2))
  addWorksheet(wb, "All_sig_genes")

  # Write data
  if (length(intersection) > 0) {
    intersection_df <- comp1_data %>% filter(gene %in% intersection) %>%
      select(gene, logFC, adj.P.Val) %>%
      rename_with(~ paste0(.x, "_", pair$label1), c(logFC, adj.P.Val)) %>%
      left_join(comp2_data %>% filter(gene %in% intersection) %>%
        select(gene, logFC, adj.P.Val) %>%
        rename_with(~ paste0(.x, "_", pair$label2), c(logFC, adj.P.Val)), by = "gene")
    writeData(wb, "Intersection", intersection_df)
  }

  if (length(unique_comp1) > 0) {
    unique_comp1_df <- comp1_data %>% filter(gene %in% unique_comp1) %>%
      select(gene, logFC, adj.P.Val)
    writeData(wb, paste0("Unique_", pair$label1), unique_comp1_df)
  }

  if (length(unique_comp2) > 0) {
    unique_comp2_df <- comp2_data %>% filter(gene %in% unique_comp2) %>%
      select(gene, logFC, adj.P.Val)
    writeData(wb, paste0("Unique_", pair$label2), unique_comp2_df)
  }

  # All significant genes summary
  all_sig_summary <- data.frame(
    Category = c(paste0("Total_sig_genes_", pair$label1),
      paste0("Total_sig_genes_", pair$label2),
      "Intersection",
      paste0("Unique_", pair$label1),
      paste0("Unique_", pair$label2)),
    Count = c(length(sig_comp1), length(sig_comp2), length(intersection),
      length(unique_comp1), length(unique_comp2))
  )
  writeData(wb, "All_sig_genes", all_sig_summary)

  # Save workbook
  xlsx_filename <- file.path(outputDataDir, paste0("venn_categories_", pair$label1, "_vs_", pair$label2, ".xlsx"))
  saveWorkbook(wb, xlsx_filename, overwrite = TRUE)

  cat("Completed analysis for:", pair$label1, "vs", pair$label2, "\n")
}
```



